---
title: "Global Refugee Flows: Baseline Visuals"
subtitle: "Who Hosts the World's Refugees? (1951–2022)"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
    self_contained: true
---

```{r setup, message=FALSE, warning=FALSE}
# ── Install any missing packages ──────────────────────────────────────────────
pkgs <- c(
  "refugees",          # UNHCR official data package (CRAN)
  "plotly",            # interactive Sankey + trend charts
  "leaflet",           # interactive map
  "dplyr", "tidyr",   # data wrangling
  "countrycode",       # ISO codes → UN regions
  "wbstats",           # World Bank income classifications
  "rnaturalearth",     # country polygons + centroids
  "rnaturalearthdata", # data dependency for rnaturalearth
  "sf",                # spatial features
  "geosphere",         # great-circle arc paths for flow lines
  "scales",            # rescale() for line widths
  "htmltools"          # HTML() for leaflet labels
)
new_pkgs <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(new_pkgs)) {
  message("Installing: ", paste(new_pkgs, collapse = ", "))
  install.packages(new_pkgs, repos = "https://cloud.r-project.org", quiet = TRUE)
}

library(refugees)
library(plotly)
library(leaflet)
library(dplyr)
library(tidyr)
library(countrycode)
library(wbstats)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(geosphere)
library(scales)
library(htmltools)
```

---

## Overview

These three baseline visuals set the stage for the DWRAP portfolio piece.
The central argument: **the vast majority of the world's displaced population
is hosted not by wealthy nations, but by developing countries** — many of which
also have comparatively liberal asylum policies on paper (as measured by DWRAP).

**Data sources:**

- **UNHCR refugee & asylum-seeker data** via the [`refugees`](https://populationstatistics.github.io/refugees/) R package (UNHCR's official package; same underlying data as the UNHCR Refugee Data Finder)
- **World Bank income classifications** via the [`wbstats`](https://CRAN.R-project.org/package=wbstats) package
- **Country polygons and centroids** via the [`rnaturalearth`](https://docs.ropensci.org/rnaturalearth/) package

---

## Data Preparation

```{r data-prep, message=FALSE, warning=FALSE}
# ── 1. UNHCR bilateral refugee data ──────────────────────────────────────────
# `refugees::population` contains country-of-origin × country-of-asylum flows
# for every year from 1951 to the most recent release.
pop <- refugees::population

flows <- pop %>%
  filter(
    !is.na(coo_iso), !is.na(coa_iso),
    coo_iso != "UNK", coa_iso != "UNK",   # drop "unknown country" records
    coo_iso != coa_iso                      # exclude same-country flows
  ) %>%
  mutate(
    # combine refugees + asylum-seekers as "total displaced in host country"
    total_displaced = coalesce(refugees, 0L) + coalesce(asylum_seekers, 0L)
  ) %>%
  filter(total_displaced > 0)

# ── 2. World Bank income groups ───────────────────────────────────────────────
wb_ctry <- wb_countries() %>%
  select(iso3c, income_level, region) %>%
  rename(iso3 = iso3c) %>%
  mutate(
    dev_status = case_when(
      income_level == "High income"                   ~ "Developed",
      income_level %in% c("Upper middle income",
                          "Lower middle income",
                          "Low income")               ~ "Developing",
      TRUE                                            ~ NA_character_
    )
  )

# ── 3. UN sub-regions via countrycode ────────────────────────────────────────
region_lkp <- codelist %>%
  select(iso3c, un.regionsub.name) %>%
  rename(iso3 = iso3c, un_subregion = un.regionsub.name) %>%
  filter(!is.na(iso3))

# ── 4. Country centroids (for flow-line anchors) ──────────────────────────────
world_sf  <- ne_countries(scale = "medium", returnclass = "sf")
centroids <- cbind(
  world_sf %>% st_drop_geometry() %>% select(iso_a3, name_long),
  world_sf %>% st_centroid() %>% st_coordinates()
) %>%
  as_tibble() %>%
  rename(iso3 = iso_a3, lng = X, lat = Y) %>%
  filter(iso3 != "-99")

# ── 5. Enrich bilateral flows ─────────────────────────────────────────────────
flows_enriched <- flows %>%
  # income group for origin
  left_join(wb_ctry %>% select(iso3, dev_status_coo = dev_status,
                                income_coo = income_level),
            by = c("coo_iso" = "iso3")) %>%
  # income group for asylum country
  left_join(wb_ctry %>% select(iso3, dev_status_coa = dev_status,
                                income_coa = income_level),
            by = c("coa_iso" = "iso3")) %>%
  # UN sub-region for origin
  left_join(region_lkp, by = c("coo_iso" = "iso3")) %>%
  mutate(
    # Simplified origin-region label for Sankey readability
    origin_region = case_when(
      un_subregion %in% c("Eastern Africa", "Western Africa",
                          "Middle Africa", "Southern Africa",
                          "Northern Africa")                     ~ "Africa",
      un_subregion %in% c("Western Asia", "Southern Asia",
                          "Central Asia")                        ~ "Middle East & S. Asia",
      un_subregion %in% c("Eastern Asia", "South-eastern Asia") ~ "East & SE Asia",
      un_subregion %in% c("Eastern Europe", "Southern Europe",
                          "Western Europe", "Northern Europe")   ~ "Europe",
      un_subregion %in% c("Central America", "South America",
                          "Caribbean")                           ~ "Latin America & Caribbean",
      un_subregion == "Northern America"                         ~ "Northern America",
      TRUE                                                        ~ "Other / Unknown"
    )
  )

cat(sprintf(
  "Records: %s  |  Years: %d–%d  |  Country pairs: %s\n",
  format(nrow(flows_enriched), big.mark = ","),
  min(flows_enriched$year),
  max(flows_enriched$year),
  format(n_distinct(paste(flows_enriched$coo_iso, flows_enriched$coa_iso)), big.mark = ",")
))
```

---

## Visual 1 — Sankey: Where Do Refugees End Up? {.tabset}

Each ribbon represents the total number of refugees + asylum-seekers flowing
from an **origin region** (left) to a **destination income group** (right) in a
given year. Ribbon width ∝ number of people. The overwhelming flow toward
*Developing* countries is the story.

> **Tip:** Change `selected_year` in the first chunk to any year in the dataset.

### Most Recent Year (2022)

```{r sankey-2022, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
build_sankey <- function(yr, title_suffix = "") {
  d <- flows_enriched %>%
    filter(year == yr, !is.na(origin_region), !is.na(dev_status_coa)) %>%
    group_by(origin_region, dev_status_coa) %>%
    summarise(total = sum(total_displaced, na.rm = TRUE), .groups = "drop") %>%
    filter(total > 0)

  if (nrow(d) == 0) return(plotly_empty() %>% layout(title = paste("No data for", yr)))

  # Fixed node order so colors are stable across tabs
  origin_order <- c("Africa", "Middle East & S. Asia", "East & SE Asia",
                    "Europe", "Latin America & Caribbean", "Northern America",
                    "Other / Unknown")
  dest_order   <- c("Developing", "Developed")

  origins_present <- intersect(origin_order, unique(d$origin_region))
  dests_present   <- intersect(dest_order,   unique(d$dev_status_coa))
  all_nodes       <- c(origins_present, dests_present)

  origin_colors <- c("#C0392B", "#E67E22", "#27AE60",
                     "#2980B9", "#8E44AD", "#16A085", "#7F8C8D")
  dest_colors   <- c("#2C3E50", "#BDC3C7")  # dark = developing, light = developed

  node_colors <- c(
    origin_colors[seq_along(origins_present)],
    dest_colors[seq_along(dests_present)]
  )

  link_df <- d %>%
    filter(origin_region %in% origins_present, dev_status_coa %in% dests_present) %>%
    mutate(
      source = match(origin_region,  all_nodes) - 1L,
      target = match(dev_status_coa, all_nodes) - 1L
    )

  plot_ly(
    type = "sankey", orientation = "h", arrangement = "snap",
    node = list(
      label     = all_nodes,
      color     = node_colors,
      pad       = 18, thickness = 22,
      line      = list(color = "white", width = 0.5)
    ),
    link = list(
      source = link_df$source,
      target = link_df$target,
      value  = link_df$total,
      color  = "rgba(180,180,180,0.30)",
      label  = paste0(
        link_df$origin_region, " → ", link_df$dev_status_coa, "<br>",
        format(link_df$total, big.mark = ","), " displaced persons"
      )
    )
  ) %>%
    layout(
      title = list(
        text = paste0(
          "<b>Refugee & Asylum-Seeker Flows by Origin Region → Destination (", yr, ")</b>",
          title_suffix,
          "<br><sup>Ribbon width = total displaced persons | Source: UNHCR via `refugees` R pkg</sup>"
        ),
        font = list(size = 13)
      ),
      font = list(size = 12),
      paper_bgcolor = "#FAFAFA"
    )
}

build_sankey(2022)
```

### 2010

```{r sankey-2010, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
build_sankey(2010)
```

### 2000

```{r sankey-2000, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
build_sankey(2000)
```

### 1990

```{r sankey-1990, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
build_sankey(1990)
```

### 1980

```{r sankey-1980, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
build_sankey(1980)
```

---

## Visual 2 — Interactive World Map: Global Hosting Burden

Countries are shaded by **total refugees + asylum-seekers hosted** (log scale,
to handle the extreme right-skew). Curved lines show the **top 40 bilateral
flows** for the selected year — line width ∝ magnitude. Hover over any country
or flow line for details.

> Change `map_year` below to explore different years.

```{r map, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
map_year <- 2022   # ← change this to any year in the dataset

# ── Totals per host country ───────────────────────────────────────────────────
hosted <- flows_enriched %>%
  filter(year == map_year) %>%
  group_by(coa_iso, coa_name) %>%
  summarise(refugees_hosted = sum(total_displaced, na.rm = TRUE), .groups = "drop")

# ── Totals per origin country ─────────────────────────────────────────────────
originated <- flows_enriched %>%
  filter(year == map_year) %>%
  group_by(coo_iso, coo_name) %>%
  summarise(refugees_from = sum(total_displaced, na.rm = TRUE), .groups = "drop")

# ── Merge onto world shapefile ────────────────────────────────────────────────
world_map <- world_sf %>%
  left_join(hosted,     by = c("iso_a3" = "coa_iso")) %>%
  left_join(originated, by = c("iso_a3" = "coo_iso")) %>%
  left_join(wb_ctry %>% select(iso3, income_level),
            by = c("iso_a3" = "iso3")) %>%
  mutate(
    refugees_hosted = replace_na(refugees_hosted, 0),
    refugees_from   = replace_na(refugees_from,   0),
    popup_text = paste0(
      "<strong>", name_long, "</strong><br>",
      "<em>", replace_na(income_level, "Income group unknown"), "</em><br>",
      "Refugees/AS hosted: <b>", format(refugees_hosted, big.mark = ","), "</b><br>",
      "Refugees/AS originating: <b>", format(refugees_from, big.mark = ","), "</b>"
    )
  )

# ── Color scale (log1p so zeros render; labels back-transform) ────────────────
pal <- colorNumeric(
  palette  = c("#FFF7EC", "#FEE8C8", "#FDD49E", "#FDBB84",
               "#FC8D59", "#EF6548", "#D7301F", "#990000"),
  domain   = log1p(world_map$refugees_hosted),
  na.color = "#D0D0D0"
)

legend_breaks <- c(0, 1e3, 1e4, 1e5, 5e5, 1e6, 2e6)
legend_labels <- c("0", "1K", "10K", "100K", "500K", "1M", "2M+")

# ── Top 40 bilateral flows ────────────────────────────────────────────────────
top_flows <- flows_enriched %>%
  filter(year == map_year) %>%
  slice_max(total_displaced, n = 40, with_ties = FALSE) %>%
  left_join(centroids %>% select(iso3, lng_coo = lng, lat_coo = lat),
            by = c("coo_iso" = "iso3")) %>%
  left_join(centroids %>% select(iso3, lng_coa = lng, lat_coa = lat),
            by = c("coa_iso" = "iso3")) %>%
  filter(!is.na(lng_coo), !is.na(lat_coo), !is.na(lng_coa), !is.na(lat_coa))

max_flow <- max(top_flows$total_displaced, na.rm = TRUE)

# ── Build leaflet map ─────────────────────────────────────────────────────────
m <- leaflet(world_map,
             options = leafletOptions(minZoom = 2, worldCopyJump = FALSE)) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = 20, lat = 15, zoom = 2) %>%
  addPolygons(
    fillColor   = ~pal(log1p(refugees_hosted)),
    weight      = 0.4, color = "white", opacity = 1,
    fillOpacity = 0.85,
    label       = ~lapply(popup_text, HTML),
    labelOptions = labelOptions(style = list("font-size" = "12px")),
    highlightOptions = highlightOptions(
      weight = 2, color = "#444", fillOpacity = 0.95, bringToFront = TRUE
    )
  ) %>%
  addLegend(
    position  = "bottomright",
    colors    = pal(log1p(legend_breaks)),
    labels    = legend_labels,
    title     = paste0("Refugees + AS Hosted<br>(", map_year, ")"),
    opacity   = 0.85
  )

# ── Overlay great-circle flow lines ──────────────────────────────────────────
for (i in seq_len(nrow(top_flows))) {
  row <- top_flows[i, ]
  lwd <- rescale(row$total_displaced, to = c(0.8, 5), from = c(0, max_flow))

  gc_path <- tryCatch(
    gcIntermediate(
      c(row$lng_coo, row$lat_coo),
      c(row$lng_coa, row$lat_coa),
      n = 60, addStartEnd = TRUE, breakAtDateLine = TRUE
    ),
    error = function(e) NULL
  )
  if (is.null(gc_path)) next

  flow_label <- HTML(paste0(
    "<strong>", row$coo_name, " → ", row$coa_name, "</strong><br>",
    format(row$total_displaced, big.mark = ","), " displaced persons"
  ))

  draw_segment <- function(seg) {
    m <<- m %>%
      addPolylines(
        lng = seg[, 1], lat = seg[, 2],
        weight = lwd, color = "#1A5276", opacity = 0.55,
        label = flow_label,
        labelOptions = labelOptions(style = list("font-size" = "11px"))
      )
  }

  if (is.list(gc_path)) {
    lapply(gc_path, draw_segment)
  } else {
    draw_segment(gc_path)
  }
}

m %>%
  addControl(
    html = HTML(paste0(
      "<div style='background:white;padding:6px 10px;border-radius:4px;",
      "font-size:11px;box-shadow:0 1px 4px rgba(0,0,0,.2);'>",
      "<strong>Top 40 bilateral flows (", map_year, ")</strong><br>",
      "Line width ∝ magnitude | Hover for details",
      "</div>"
    )),
    position = "topright"
  )
```

---

## Visual 3 — Trend: Share of Refugees in Developing Countries, 1951–2022

The simplest framing of the core argument: what **share** of all refugees
globally is hosted in developing (non-high-income) countries, year by year?

```{r trend, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
trend <- flows_enriched %>%
  filter(!is.na(dev_status_coa)) %>%
  group_by(year, dev_status_coa) %>%
  summarise(total = sum(total_displaced, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  mutate(
    grand_total = sum(total),
    share       = total / grand_total
  ) %>%
  ungroup()

dev_share <- trend %>% filter(dev_status_coa == "Developing")

# Global totals (both categories) for the secondary axis context
totals_by_year <- trend %>%
  group_by(year) %>%
  summarise(grand_total = sum(total), .groups = "drop")

plot_ly() %>%
  # Area fill under the share line
  add_trace(
    data = dev_share,
    x = ~year, y = ~share,
    type = "scatter", mode = "none",
    fill = "tozeroy", fillcolor = "rgba(192,57,43,0.12)",
    showlegend = FALSE, hoverinfo = "none"
  ) %>%
  # Share line (primary y-axis)
  add_trace(
    data = dev_share,
    x = ~year, y = ~share,
    type = "scatter", mode = "lines+markers",
    name = "Share in developing countries",
    line   = list(color = "#C0392B", width = 2.5),
    marker = list(color = "#C0392B", size = 4),
    hovertemplate = paste0(
      "<b>%{x}</b><br>",
      "Share in developing countries: <b>%{y:.1%}</b>",
      "<extra></extra>"
    )
  ) %>%
  # 50% reference line annotation
  add_segments(
    x = ~min(dev_share$year), xend = ~max(dev_share$year),
    y = 0.5, yend = 0.5,
    line = list(color = "gray", dash = "dot", width = 1),
    showlegend = FALSE, hoverinfo = "none"
  ) %>%
  layout(
    title = list(
      text = paste0(
        "<b>Share of World's Refugees Hosted by Developing Countries</b><br>",
        "<sup>1951–2022 | Refugees + asylum-seekers | Source: UNHCR via `refugees` R package</sup>"
      ),
      font = list(size = 13)
    ),
    xaxis = list(
      title = "Year", dtick = 10,
      showgrid = TRUE, gridcolor = "#EEEEEE"
    ),
    yaxis = list(
      title = "Share of global displaced population",
      tickformat = ".0%", range = c(0, 1),
      showgrid = TRUE, gridcolor = "#EEEEEE"
    ),
    annotations = list(list(
      x = min(dev_share$year) + 2, y = 0.52,
      text = "50% threshold", showarrow = FALSE,
      font = list(color = "gray", size = 10)
    )),
    paper_bgcolor = "#FAFAFA",
    plot_bgcolor  = "#FAFAFA",
    legend = list(x = 0.02, y = 0.15)
  )
```

---

*Next step: merge these baseline visuals with the DWRAP policy index data
(`dwrap_1951_2022.dta`) to show that developing countries with high hosting
burdens often also score higher on de jure policy liberality.*

*Data: UNHCR Population Statistics via the `refugees` R package (CRAN).
Income classifications: World Bank Development Indicators. This document is
part of a larger portfolio piece on the
[DWRAP dataset](https://datacatalog.worldbank.org/search/dataset/0064424).*
