---
title: "DWRAP — Policy Liberality Trends: Global North vs. Global South"
subtitle: "Step 2 of the Narrative: The Global South Is Liberalizing"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
    self_contained: true
---

```{r setup, message=FALSE, warning=FALSE}
pkgs <- c("haven", "dplyr", "tidyr", "ggplot2", "plotly",
          "countrycode", "wbstats", "scales", "ggrepel", "forcats")
new_pkgs <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(new_pkgs)) install.packages(new_pkgs, repos = "https://cloud.r-project.org", quiet = TRUE)

library(haven)
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(countrycode)
library(wbstats)
library(scales)
library(ggrepel)
```

---

## Overview

This document produces **Visual 4** of the DWRAP portfolio narrative:

> *As the world's refugee burden has shifted toward developing countries,
> those same countries have — on paper — been **liberalizing** their asylum
> and refugee policies at a faster rate than wealthy nations.*

The chart shows individual country policy trajectories as light background
traces, with two bold trend lines distinguishing the **Global North** from
the **Global South**.

---

## A Note on "Global North" vs. "Global South"

This is a contested classification. Three operationalizations are common:

| Method | Definition of "Global North" | Pros | Cons |
|--------|------------------------------|------|------|
| **World Bank income group** (used here as primary) | High-income countries | Objective, data-driven, annually updated | Conflates economic status with geopolitical category |
| **OECD membership** | 38 current OECD members | Politically meaningful "developed world" proxy | Includes Turkey, Mexico, Chile; excludes Singapore, Saudi Arabia |
| **UN development status / HDI** | Very High / High HDI | Reflects human development broadly | Changes over time, less binary |

**This file implements both the World Bank and OECD methods** and plots them
side by side so you can choose which framing best fits the narrative.

---

## 1. Data Loading & Exploration

```{r load-data, message=FALSE, warning=FALSE}
# Path is relative to this Rmd's location (DWRAP/ subfolder)
dwrap_path <- "../dwrap_1951_2022.dta"

if (!file.exists(dwrap_path)) {
  stop(paste(
    "Data file not found at:", dwrap_path,
    "\nUpdate the path to point to dwrap_1951_2022.dta"
  ))
}

dwrap_raw <- read_dta(dwrap_path)

# ── Explore structure ─────────────────────────────────────────────────────────
cat("Dimensions:", nrow(dwrap_raw), "rows ×", ncol(dwrap_raw), "columns\n\n")
cat("All column names:\n")
print(names(dwrap_raw))

cat("\nYear range:", min(dwrap_raw$year, na.rm = TRUE),
    "–", max(dwrap_raw$year, na.rm = TRUE), "\n")
cat("Unique countries:", n_distinct(dwrap_raw$country), "\n")
```

```{r identify-index-column, message=FALSE, warning=FALSE}
# ── Identify the overall policy index column ──────────────────────────────────
# Based on the DWRAP tech guide, the overall policy score:
#   - ranges from 0 to ~0.674 (mean ≈ 0.167)
#   - is numeric / double
#
# Run this chunk to see candidate columns — then set `policy_col` below.

numeric_cols <- dwrap_raw %>%
  select(where(is.numeric)) %>%
  select(-year) %>%
  summarise(across(everything(), list(
    min  = ~min(., na.rm = TRUE),
    max  = ~max(., na.rm = TRUE),
    mean = ~mean(., na.rm = TRUE)
  ))) %>%
  pivot_longer(everything(),
               names_to = c("variable", ".value"),
               names_pattern = "(.+)_(min|max|mean)") %>%
  filter(min >= 0, max <= 1, mean > 0.05, mean < 0.5) %>%
  arrange(desc(mean))

cat("Candidate policy index columns (0–1 range, plausible mean):\n")
print(numeric_cols)

# ── SET THIS after reviewing the output above ─────────────────────────────────
#
# Common names seen in DWRAP releases: "index", "policy_index", "i_policy",
# "lib_index", "overall_index". Set the correct name below.
#
policy_col  <- "index"       # ← UPDATE if column name differs
access_col  <- "access"      # ← field-level index (Access)
services_col <- "services"   # ← field-level index (Services)
# (remaining fields: livelihoods, movement, participation)
```

---

## 2. North / South Classification

```{r classify, message=FALSE, warning=FALSE}
# ── World Bank income groups ──────────────────────────────────────────────────
wb_ctry <- wb_countries() %>%
  select(iso3c, country = country, income_level) %>%
  mutate(
    north_wb = income_level == "High income",
    ns_wb    = if_else(north_wb, "Global North\n(High Income)", "Global South\n(Non-High Income)")
  )

# ── OECD membership (current 38 members as of 2024) ──────────────────────────
oecd_iso3 <- c(
  "AUS","AUT","BEL","CAN","CHL","COL","CRI","CZE","DNK","EST",
  "FIN","FRA","DEU","GRC","HUN","ISL","IRL","ISR","ITA","JPN",
  "KOR","LVA","LTU","LUX","MEX","NLD","NZL","NOR","POL","PRT",
  "SVK","SVN","ESP","SWE","CHE","TUR","GBR","USA"
)

oecd_df <- tibble(
  iso3c   = oecd_iso3,
  north_oecd = TRUE,
  ns_oecd    = "Global North\n(OECD)"
)

# ── Add ISO3 codes to DWRAP using countrycode ─────────────────────────────────
# DWRAP uses country names; we convert to ISO3 for joining.
dwrap <- dwrap_raw %>%
  mutate(
    iso3 = countrycode(country, origin = "country.name", destination = "iso3c",
                       warn = FALSE)
  )

# Warn about any countries that didn't match
unmatched <- dwrap %>%
  filter(is.na(iso3)) %>%
  distinct(country)
if (nrow(unmatched) > 0) {
  message("Countries without ISO3 match (", nrow(unmatched), "):")
  print(unmatched$country)
}

# ── Merge classifications ─────────────────────────────────────────────────────
dwrap <- dwrap %>%
  left_join(wb_ctry %>% select(iso3c, north_wb, ns_wb),
            by = c("iso3" = "iso3c")) %>%
  left_join(oecd_df,
            by = c("iso3" = "iso3c")) %>%
  mutate(
    north_oecd = replace_na(north_oecd, FALSE),
    ns_oecd    = if_else(north_oecd, "Global North\n(OECD)", "Global South\n(Non-OECD)")
  )

cat("World Bank classification:\n")
print(table(dwrap$ns_wb, useNA = "always"))
cat("\nOECD classification:\n")
print(table(dwrap$ns_oecd, useNA = "always"))
```

---

## 3. Aggregate to Country-Year Means

```{r aggregate, message=FALSE, warning=FALSE}
# If the data has one row per law (multiple rows per country-year),
# collapse to country-year by taking the most recent law's score.
# If already country-year panel, this is a no-op.

# Check: rows per country-year
rows_per_cy <- dwrap %>%
  count(country, year) %>%
  summarise(max_rows = max(n), mean_rows = mean(n))
cat("Max rows per country-year:", rows_per_cy$max_rows, "\n")
cat("Mean rows per country-year:", round(rows_per_cy$mean_rows, 2), "\n")

# Collapse to country-year (take max index = most liberal law if multiple)
# Adjust `policy_col` at the top if the column name differs.
cy_panel <- dwrap %>%
  rename(policy_index = all_of(policy_col)) %>%
  group_by(country, iso3, year, ns_wb, ns_oecd, north_wb, north_oecd) %>%
  summarise(
    policy_index = max(policy_index, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.infinite(policy_index), !is.na(policy_index))

cat("\nCountry-year panel:", nrow(cy_panel), "rows\n")
cat("Countries:", n_distinct(cy_panel$country), "\n")
```

---

## 4. Visual 4 — Policy Liberality Trends: North vs. South

### Version A — World Bank Income Classification

```{r plot-wb, message=FALSE, warning=FALSE, fig.width=11, fig.height=7}
# ── Group averages ────────────────────────────────────────────────────────────
avg_wb <- cy_panel %>%
  filter(!is.na(ns_wb)) %>%
  group_by(year, ns_wb) %>%
  summarise(mean_index = mean(policy_index, na.rm = TRUE),
            n = n(), .groups = "drop")

# ── Palette ───────────────────────────────────────────────────────────────────
palette_wb <- c(
  "Global North\n(High Income)"     = "#2166AC",   # blue
  "Global South\n(Non-High Income)" = "#D6604D"    # red-orange
)

# ── Placeholder annotations: add your key events here ───────────────────────
# Format: list(year, label, y_nudge)
# y_nudge shifts label up/down from the group mean to avoid overlap.
events <- tribble(
  ~year, ~label,                          ~y_pos,
  1951,  "1951 Refugee\nConvention",       0.62,
  1967,  "1967\nProtocol",                 0.62,
  1969,  "OAU\nConvention",                0.55,
  1984,  "Cartagena\nDeclaration",         0.55,
  2018,  "Global Compact\non Refugees",    0.62
  # ↑ Add or edit events here — x = year, y_pos = height on 0–1 scale
)

# ── ggplot base ───────────────────────────────────────────────────────────────
p_wb <- ggplot() +
  # Individual country traces (light gray background)
  geom_line(
    data = cy_panel %>% filter(!is.na(ns_wb)),
    aes(x = year, y = policy_index, group = country),
    color = "#CCCCCC", linewidth = 0.25, alpha = 0.5
  ) +
  # Group average lines
  geom_line(
    data = avg_wb,
    aes(x = year, y = mean_index, color = ns_wb),
    linewidth = 1.8
  ) +
  # International legal standard reference line
  # (≈ 0.30 based on DWRAP tech guide Figure 5; adjust if needed)
  geom_hline(yintercept = 0.30, linetype = "dashed",
             color = "#888888", linewidth = 0.6) +
  annotate("text", x = 1953, y = 0.315, label = "Int'l legal standard",
           size = 2.8, color = "#888888", hjust = 0) +
  # Event annotation lines
  geom_vline(data = events, aes(xintercept = year),
             linetype = "dotted", color = "#AAAAAA", linewidth = 0.5) +
  # Event labels
  geom_text(data = events,
            aes(x = year, y = y_pos, label = label),
            size = 2.5, color = "#555555", hjust = -0.08,
            lineheight = 0.85, fontface = "italic") +
  # Scales & theme
  scale_color_manual(values = palette_wb, name = NULL) +
  scale_x_continuous(breaks = seq(1950, 2020, 10), expand = c(0.01, 0)) +
  scale_y_continuous(limits = c(0, 0.70), labels = label_number(accuracy = 0.01)) +
  labs(
    title    = "Asylum Policy Liberality Over Time: Global North vs. Global South",
    subtitle = paste0(
      "North = World Bank High-Income countries (n = ",
      n_distinct(filter(cy_panel, north_wb)$country),
      "); South = all others (n = ",
      n_distinct(filter(cy_panel, !replace_na(north_wb, FALSE))$country),
      ")\n",
      "Light gray = individual country trends  |  Bold lines = group means  |  Index scaled 0–1"
    ),
    x = NULL, y = "Policy Liberality Index (0–1)",
    caption = "Source: DWRAP (Blair, Grossman & Weinstein / World Bank, 2024). Income classifications: World Bank."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = c(0.12, 0.88),
    legend.text      = element_text(size = 10, lineheight = 0.9),
    legend.key.width = unit(1.5, "cm"),
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 10, color = "#555555"),
    plot.caption     = element_text(size = 8,  color = "#888888"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

# Static version
p_wb

# Interactive version (uncomment to use instead)
# ggplotly(p_wb, tooltip = c("x", "y", "colour")) %>%
#   layout(legend = list(x = 0.02, y = 0.95))
```

### Version B — OECD Membership Classification

```{r plot-oecd, message=FALSE, warning=FALSE, fig.width=11, fig.height=7}
avg_oecd <- cy_panel %>%
  filter(!is.na(ns_oecd)) %>%
  group_by(year, ns_oecd) %>%
  summarise(mean_index = mean(policy_index, na.rm = TRUE),
            n = n(), .groups = "drop")

palette_oecd <- c(
  "Global North\n(OECD)"     = "#1B7837",   # green
  "Global South\n(Non-OECD)" = "#E08214"    # amber
)

p_oecd <- ggplot() +
  geom_line(
    data = cy_panel %>% filter(!is.na(ns_oecd)),
    aes(x = year, y = policy_index, group = country),
    color = "#CCCCCC", linewidth = 0.25, alpha = 0.5
  ) +
  geom_line(
    data = avg_oecd,
    aes(x = year, y = mean_index, color = ns_oecd),
    linewidth = 1.8
  ) +
  geom_hline(yintercept = 0.30, linetype = "dashed",
             color = "#888888", linewidth = 0.6) +
  annotate("text", x = 1953, y = 0.315, label = "Int'l legal standard",
           size = 2.8, color = "#888888", hjust = 0) +
  geom_vline(data = events, aes(xintercept = year),
             linetype = "dotted", color = "#AAAAAA", linewidth = 0.5) +
  geom_text(data = events,
            aes(x = year, y = y_pos, label = label),
            size = 2.5, color = "#555555", hjust = -0.08,
            lineheight = 0.85, fontface = "italic") +
  scale_color_manual(values = palette_oecd, name = NULL) +
  scale_x_continuous(breaks = seq(1950, 2020, 10), expand = c(0.01, 0)) +
  scale_y_continuous(limits = c(0, 0.70), labels = label_number(accuracy = 0.01)) +
  labs(
    title    = "Asylum Policy Liberality Over Time: Global North vs. Global South",
    subtitle = paste0(
      "North = OECD member states (n = ", length(oecd_iso3), "); ",
      "South = non-OECD countries\n",
      "Light gray = individual country trends  |  Bold lines = group means  |  Index scaled 0–1"
    ),
    x = NULL, y = "Policy Liberality Index (0–1)",
    caption = "Source: DWRAP (Blair, Grossman & Weinstein / World Bank, 2024). OECD membership as of 2024."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = c(0.12, 0.88),
    legend.text      = element_text(size = 10, lineheight = 0.9),
    legend.key.width = unit(1.5, "cm"),
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 10, color = "#555555"),
    plot.caption     = element_text(size = 8,  color = "#888888"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

p_oecd
```

### Version C — Interactive (plotly), World Bank Classification

Use this version for embedding in the portfolio website — it supports hover,
zoom, and pan.

```{r plot-interactive, message=FALSE, warning=FALSE, fig.width=11, fig.height=7}
# Build a plotly version with country-level hover detail
build_interactive_trend <- function(ns_col, palette, title_note) {
  cy_filtered <- cy_panel %>% filter(!is.na(.data[[ns_col]]))
  avg         <- cy_filtered %>%
    group_by(year, group = .data[[ns_col]]) %>%
    summarise(mean_index = mean(policy_index, na.rm = TRUE), .groups = "drop")

  fig <- plot_ly()

  # Individual country traces (gray, no legend)
  countries <- unique(cy_filtered$country)
  for (cty in countries) {
    d <- cy_filtered %>% filter(country == cty)
    fig <- fig %>%
      add_trace(
        data = d, x = ~year, y = ~policy_index,
        type = "scatter", mode = "lines",
        line = list(color = "rgba(180,180,180,0.35)", width = 0.6),
        showlegend = FALSE,
        hoverinfo = "text",
        text = ~paste0(country, " (", year, "): ", round(policy_index, 3))
      )
  }

  # Group average traces
  for (grp in unique(avg$group)) {
    d   <- avg %>% filter(group == grp)
    col <- palette[grp]
    # Clean label for legend (remove \n)
    lbl <- gsub("\n", " ", grp)
    fig <- fig %>%
      add_trace(
        data = d, x = ~year, y = ~mean_index,
        type = "scatter", mode = "lines",
        name = lbl,
        line = list(color = col, width = 3),
        hovertemplate = paste0(lbl, " | %{x}: %{y:.3f}<extra></extra>")
      )
  }

  # Event annotations
  annotations <- lapply(seq_len(nrow(events)), function(i) {
    list(
      x = events$year[i], y = events$y_pos[i],
      text = gsub("\n", "<br>", events$label[i]),
      showarrow = TRUE, arrowhead = 0, arrowwidth = 1,
      arrowcolor = "#AAAAAA", ax = 20, ay = -25,
      font = list(size = 9, color = "#555555"),
      xref = "x", yref = "y"
    )
  })

  # Int'l legal standard line
  shapes <- list(list(
    type = "line",
    x0 = min(cy_panel$year), x1 = max(cy_panel$year),
    y0 = 0.30, y1 = 0.30,
    line = list(color = "#AAAAAA", dash = "dot", width = 1)
  ))

  fig %>%
    layout(
      title = list(
        text = paste0(
          "<b>Asylum Policy Liberality Over Time: Global North vs. Global South</b><br>",
          "<sup>", title_note, " | Light traces = individual countries | Hover for country detail</sup>"
        ),
        font = list(size = 13)
      ),
      xaxis = list(title = "Year", dtick = 10,
                   showgrid = TRUE, gridcolor = "#EEEEEE"),
      yaxis = list(title = "Policy Liberality Index (0–1)",
                   range = c(0, 0.72),
                   showgrid = TRUE, gridcolor = "#EEEEEE"),
      annotations = annotations,
      shapes      = shapes,
      legend      = list(x = 0.02, y = 0.95, bgcolor = "rgba(255,255,255,0.8)"),
      paper_bgcolor = "#FAFAFA",
      plot_bgcolor  = "#FAFAFA",
      hovermode = "closest"
    )
}

build_interactive_trend(
  ns_col     = "ns_wb",
  palette    = c("Global North\n(High Income)"     = "#2166AC",
                 "Global South\n(Non-High Income)" = "#D6604D"),
  title_note = "North = World Bank High Income; South = all others"
)
```

---

## 5. Interpretation Notes

Key things to look for in the chart:

- **Convergence or divergence**: Is the Global South trend line rising faster
  than the Global North line, narrowing the gap over time?
- **Inflection points**: Where do the trends spike or flatten? These are
  candidate moments for annotation — you will identify the qualitative drivers.
- **Below the legal standard**: Note years when both group means sit *below*
  the international legal standard dashed line. This is a key talking point.
- **Variance within groups**: The spread of gray lines tells its own story —
  there is enormous heterogeneity *within* the Global South.

---

*Next step: Merge this policy index trend with refugee hosting data to show
whether countries that host more refugees also adopt more liberal policies —
and whether that relationship holds differently in the North vs. South.*
